---
title: "Fish Tissue Contamination and its Effects on Human and Ecosystem Health"
subtitle: "An examination of trends in fish tissue composition in the Great Lakes"
author: "Jaclyn Narleski and Victoria Thompson"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
---
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(here)
library(lubridate)
library(tidyverse)
library(agricolae)
library(ggplot2)

# Check working directory
here()

# Set your ggplot theme
laketheme <- theme(
  panel.background = element_rect(fill = "gray90"),
  panel.border = element_rect(color = "darkcyan", fill = NA, linetype = 1),
  plot.title = element_text(
    hjust = 0.5, size = 12, color = "darkcyan", 
    face = "bold"),
  axis.title = element_text(size = 10, color = "darkcyan"),
  panel.grid.major = element_line(
    color = "aquamarine4", linetype = 2, size = 0.25),
  panel.grid.minor = element_line(
    color = "aquamarine3", linetype = 3, size = 0.25),
  axis.text = element_text(color = "aquamarine4"),
  axis.ticks = element_line(color = "aquamarine4"),
  legend.title = element_text(size = 8, color = "darkcyan"),
  legend.position = "right"
)

# Set theme
theme_set(laketheme)

# Load your datasets

# 2010 data
fattyacids_2010_raw <- read.csv("RawData/2010_FattyAcids_Raw.csv")
mercury_2010_raw <- read.csv("RawData/2010_Mercury_Raw.csv")
PCB_2010_raw <- read.csv("RawData/2010_PCBs_Raw.csv")
PFA_2010_raw <- read.csv("RawData/2010_PFAs_Raw.csv")

# 2015 data
fattyacids_2015_raw <- read.csv("RawData/2015_FattyAcids_Raw.csv")
mercury_2015_raw <- read.csv("RawData/2015_Mercury_Raw.csv")
PCB_2015_raw <- read.csv("RawData/2015_PCBs_Raw.csv")
PFA_2015_raw <- read.csv("RawData/2015_PFAs_Raw.csv")

# 2020 data
fattyacids_2020_raw <- read.csv("RawData/2020_FattyAcids_Raw.csv")
mercury_2020_raw <- read.csv("RawData/2020_Mercury_Raw.csv")
PCB_2020_raw <- read.csv("RawData/2020_PCBs_Raw.csv")
PFA_2020_raw <- read.csv("RawData/2020_PFAs_Raw.csv")
```

# Rationale and Research Questions
The Great Lakes, which collectively represent the largest surface freshwater system on Earth, are critical ecological and economic resources for millions of people. These lakes support diverse aquatic life, serve as a source of drinking water, provide recreational opportunities, and underpin industries such as fishing. However, over the past century, the Great Lakes have been threatened by environmental stressors, including pollution from industrial waste, agricultural runoff, and urban development.[1] The contaminants affecting these ecosystems pose serious risks to both aquatic life and human health.

Fish accumulate pollutants in their tissues, decreasing their health and altering ecological dynamics within the lakes. In 2010, the EPA launched the NCCA (National Coastal Condition Assessment) to “assess the condition of nearshore waters along the marine and Great Lakes coasts of the United States and to evaluate changes affecting the quality of these waters over time”.[2] This project includes sampling and testing fish tissues for particularly harmful long-lasting chemicals, including mercury, PFAS, PCBs, and fatty acids. The EPA hopes that data on these contamination levels will allow them to be better informed on the current risks to fish, and the aquatic and human lives that depend on them.

Between the fragile ecosystems that depend on fish to the entire human populations to rely on them, studying contamination is of the utmost importance. This has led to the development of the following research questions to guide our investigation:

<ol><b>
  <li>How does fish tissue contamination vary within and between each lake?</li>
  <li>How do contaminants vary over time in each lake and overall?</li>
  <li>Are different species more vulnerable to different contaminants?</li>
</b></ol>

\newpage

# Dataset Information
The datasets used in this analysis were sourced from the Environmental Protection Agency (EPA) website. They were selected from available National Coastal Condition Assessments (NCCA), which are surveys designed to assess long-term conditions of nearshore waters along the coasts of the Great Lakes. Our data specifically involves the data collected from Fish Fillet Tissue studies, where samples of fish muscle from commonly caught recreational fish species are taken and analyzed for mercury, PFAs (per- & polyfluoroalkl substances), PCB (polychlorinated biphenyls) Congeners & Aroclors, and Omega-3 & Omega-6 fatty acids. The EPA uses the results from these studies to evaluate the water quality of the Great Lakes and more specifically, determine the health impacts humans may have from consuming fish caught in the Great Lakes.

Fish contaminant data is collected using the following procedures, per the 2020 NCCA Field Operations Manual:
- Sites are randomly selected along U.S. coasts of the Great Lakes
- Crews collect fish at all NCCA sites, yielding a minimum total of 300g total mass per target species
- Fish are identified, measured, labelled, frozen, and brough to a lab
- Specimens are cleaned to prevent contamination with non-marine chemicals
- Fish are filleted (scales are removed, but skin and belly are left intact)
- A variety of lab procedures are performed in accordance with EPA guidelines to determine the lipid content, concentrations of metals, pesticides, and PCBs in each fish

The Environmental Protection Agency has conducted periodic studies over the past 15 years assessing the coastal conditions of marine and the Great Lakes shores. We focus on data from 2010, 2015, and 2020. The basic structure of each is described below (Table 3.1). 

**Table 3.1. Raw Dataset Column Descriptions**
<br>
<p style="color: lightgray;">*adapted from 2020 GLHHFTS Fish Tissue Data Dictionary for Mercury, PCBs, and PFAS Results November 2023*.</p>

| Column Name | Description |
|:----|:------|
| EPA Region | The EPA Region in which the sample was collected | 
| State |  2-letter abbreviation for the state in which the sample was collected | 
| Lake | The Great Lake the sample is from |
| Site ID | "NGL" + site selection year + state + 4-digit site location |
| EPA Sample ID | 6-digit code assigned by the EPA |
| Tissue Type | The type of fish tissue used for analysis; we use only fillet data |
| % Lipids | The lipid content of the sample, measured independently of the chemical analysis |
| Method | EPA-standardized method of chemical analysis |
| Analyte | The chemical of interest in each dataset |
| Amount | Amount of chemcial detected in the sample |

**Dataset Wrangling Procedures**
<br>
Each of the 5-year publications from the EPA contains 4 separate data files arranged by contaminant (Mercury, PFAS, PCBs, Fatty Acids). There are also 3 metadata files and 3 quality assurance reports for each. [3] Each of the 4 separate contaminant data files were loaded into R as raw .xlsx files. The columns from Table 3.1 were selected from the raw file, and a new column was added for the collection year. This was repeated for all 12 datasets (4 contaminants for each of 3 years). There are slight variations in the dataset formatting between years, leading to differences in the pipe function to account for them. For example, some columns are renamed as to all have a consistent naming conventions. Each wrangled dataset was reassigned to a "processed" .csv file for analysis.
<br>
<br>
**Summary of Data Structure**
<br>
The basic structure of each of the raw datasets can be found in Table 3.1. The specific variables being investigated from  each of the processed files can be found below (Table 3.2).
<br>
<br>
**Table 3.2. Processed Dataset Column Descriptions**

| Variable | Description |
|:----|:------|
| State |  2-letter abbreviation for the state in which the sample was collected | 
| Lake | The Great Lake the sample is from |
| Scientific.Name | Taxonomic name of fish sampled |
| Common.Name | Common name of fish sampled |
| Family | Taxonomic family of fish sampled |
| Niche | Codified ecological niche of the fish sampled; for example, "BD" is" "Benthic-Demersal" |
| Method | EPA-standardized method of chemical analysis |
| Analyte | The chemical of interest in each dataset |
| Amount | Amount of chemcial detected in the sample |
| Unit.1 | The original unit the chemical was measured in. See Table 3.3 |
| Unit.2 | The amount of chemical converted to common units. Only for some contaminants. See Table 3.3 |
<br>

**Table 3.3. Contaminant Units**

| Contaminant | Unit 1 (Original) | Unit 2 (Converted)|
|:----|:------|:------|
| Mercury | ng/g (nanograms per gram)  | ppb (parts per billion)|
| Fatty Acids | mg/g (millograms per gram)| N/A |
| PCBs | ng/g (nanograms per gram)  | ppb (parts per billion)|
| PFAs  | ng/g (nanograms per gram)  | ppb (parts per billion)|


```{r wrangling, include=FALSE}}

# Wrangle mercury data and write in as csvs

mercury_2010_processed <- mercury_2010_raw %>%
  select(State, Lake.Name, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  mutate(Year = rep(2010, nrow(mercury_2010_raw))) %>%
  rename(Lake = Lake.Name)

write.csv(mercury_2010_processed, 
"~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2010_Mercury_Processed.csv")

mercury_2015_processed <- mercury_2015_raw %>%
  select(State, Lake, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  mutate(Year = rep(2015, nrow(mercury_2015_raw)))

write.csv(mercury_2015_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2015_Mercury_Processed.csv")

mercury_2020_processed <- mercury_2020_raw %>%
  select(State, Lake, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Units.1, Units.2) %>%
  mutate(Year = rep(2020, nrow(mercury_2020_raw)))

write.csv(mercury_2020_processed, 
"~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2020_Mercury_Processed.csv")

# Wrangle PFAs data and write in as csvs

PFA_2010_processed <- PFA_2010_raw %>%
  select(State, Lake.Name, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  mutate(Year = rep(2010, nrow(PFA_2010_raw))) %>%
  rename(Lake = Lake.Name)

write.csv(PFA_2010_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2010_PFAs_Processed.csv")

PFA_2015_processed <- PFA_2015_raw %>%
  select(State, Lake, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  mutate(Year = rep(2015, nrow(PFA_2015_raw)))

write.csv(PFA_2015_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2015_PFAs_Processed.csv")

PFA_2020_processed <- PFA_2020_raw %>%
  select(State, Lake, Scientific.Name, Common.Name, Family, Niche,
         Chemical, Amount, Units.1, Units.2) %>%
  mutate(Year = rep(2020, nrow(PFA_2020_raw)))

write.csv(PFA_2020_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2020_PFAs_Processed.csv")

# Wrangle PCB data and write in as csvs

PCB_2010_processed <- PCB_2010_raw %>%
  select(State, Lake.Name, Species.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  filter(Analyte == "Total PCBs") %>%
  mutate(Year = 2010) %>%
  rename(Lake = Lake.Name,
         Scientific.Name = Species.Name)

write.csv(PCB_2010_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2010_PCBs_Processed.csv")

PCB_2015_processed <- PCB_2015_raw %>%
  select(State, Lake, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  filter(Analyte == "Total PCBs") %>%
  mutate(Year = 2015)

write.csv(PCB_2015_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2015_PCBs_Processed.csv")

PCB_2020_processed <- PCB_2020_raw %>%
  select(State, Lake, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Units.1, Units.2) %>%
  filter(Analyte == "Total PCBs") %>%
  mutate(Year = 2020)

write.csv(PCB_2020_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2020_PCBs_Processed.csv")

# Wrangle fatty acid data and write in as csvs

fattyacids_2010_processed <- fattyacids_2010_raw %>%
   select(State, Lake.Name, Species.ID, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Abbreviation, Amount, Units) %>%
  mutate(Year = 2010, Amount = coalesce(Amount, 0)) %>%
  rename(Lake = Lake.Name)

write.csv(fattyacids_2010_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2010_FattyAcids_Processed.csv")

fattyacids_2015_processed <- fattyacids_2015_raw %>%
   select(State, Lake, Species.ID, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Unit.1, Unit.2) %>%
  mutate(Year = 2015, Amount = coalesce(Amount, 0))

write.csv(fattyacids_2015_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2015_FattyAcids_Processed.csv")

fattyacids_2020_processed <- fattyacids_2020_raw %>%
   select(State, Lake, Species.ID, Scientific.Name, Common.Name, Family, Niche,
         Analyte, Amount, Units) %>%
  mutate(Year = 2020, Amount = coalesce(Amount, 0))

write.csv(fattyacids_2020_processed, "~/FinalProject/NarleskiThompson_ENV872_EDA_FinalProject/ProcessedData/2020_FattyAcids_Processed.csv")

```

\newpage

# Exploratory Analysis
Before beginning analysis aimed at answering our research questions, we explore it in several ways. First, to orient ourselves spatially, we mapped the all of the sites sampled in 2020. We aren't able use specific site location in our analysis, due to a lack of spatial data in the 2010 and 2015 datasets. However, it is noted in the metadata (from which we etxract the sites and their coordinates) that many of the 2020 sites were locations revisited from previous sample years (Figure 4.1).[4]

```{r}
sampling_sites <- read.csv("RawData/2020sitelocations.csv")

ggplot(data = sampling_sites , aes(x = LONGITUDE., y = lat)) +
  geom_point(color = "blue") +
  labs(title = "Fish Monitoring Sites", x = "Longitude", y = "Latitude") +
  theme_minimal()

#Insert exploratory visualizations of your dataset. This may include, but is not limited to, graphs illustrating the distributions of variables of interest and/or maps of the spatial context of your dataset. Format your R chunks so that graphs are displayed but code is not displayed. Accompany these graphs with text sections that describe the visualizations and provide context for further analyses.

#Each figure should be accompanied by a caption, and each figure should be referenced within the text.

#Scope: think about what information someone might want to know about the dataset before analyzing it statistically. How might you visualize this information?
```


```{r}

# Mercury exploratory analysis

# 2010
Mercury_bylake_2010 <- mercury_2010_processed %>%
  group_by(Lake, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

Mercury_byfish_2010 <- mercury_2010_processed %>%
  group_by(Scientific.Name, Common.Name, Family, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

Mercury_byniche_2010 <- mercury_2010_processed %>%
  group_by(Niche, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

# 2015
Mercury_bylake_2015 <- mercury_2015_processed %>%
  group_by(Lake, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

Mercury_byfish_2015 <- mercury_2015_processed %>%
  group_by(Scientific.Name, Common.Name, Family, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

Mercury_byniche_2015 <- mercury_2015_processed %>%
  group_by(Niche, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

# 2020
Mercury_bylake_2020 <- mercury_2020_processed %>%
  group_by(Lake, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

Mercury_byfish_2020 <- mercury_2020_processed %>%
  group_by(Scientific.Name, Common.Name, Family, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

Mercury_byniche_2020 <- mercury_2020_processed %>%
  group_by(Niche, Year) %>%
  summarise(Mean.Mercury = mean(Amount))

# Combined by lake 
Mercury_bylake_combined <- 
  rbind(Mercury_bylake_2010, Mercury_bylake_2015, Mercury_bylake_2020)

# Combined by fish species
Mercury_byfish_combined <-
  rbind(Mercury_byfish_2010, Mercury_byfish_2015, Mercury_byfish_2020)

# Combined by niche
Mercury_byniche_combined <-
  rbind(Mercury_byniche_2010, Mercury_byniche_2015, Mercury_byniche_2020)

```
```{r}

# PCB exploratory analysis

# 2010
PCB_bylake_2010 <- PCB_2010_processed %>%
  group_by(Lake, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

PCB_byfish_2010 <- PCB_2010_processed %>%
  group_by(Scientific.Name, Common.Name, Family, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

PCB_byniche_2010 <- PCB_2010_processed %>%
  group_by(Niche, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

# 2015
PCB_bylake_2015 <- PCB_2015_processed %>%
  group_by(Lake, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

PCB_byfish_2015 <- PCB_2015_processed %>%
  group_by(Scientific.Name, Common.Name, Family, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

PCB_byniche_2015 <- PCB_2015_processed %>%
  group_by(Niche, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

# 2020
PCB_bylake_2020 <- PCB_2020_processed %>%
  group_by(Lake, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

PCB_byfish_2020 <- PCB_2020_processed %>%
  group_by(Scientific.Name, Common.Name, Family, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

PCB_byniche_2020 <- PCB_2020_processed %>%
  group_by(Niche, Year) %>%
  summarise(Mean.PCBs = mean(Amount))

# Combined by lake 
PCB_bylake_combined <- 
  rbind(PCB_bylake_2010, PCB_bylake_2015, PCB_bylake_2020)

# Combined by fish species
PCB_byfish_combined <-
  rbind(PCB_byfish_2010, PCB_byfish_2015, PCB_byfish_2020)

# Combined by niche
PCB_byniche_combined <-
  rbind(PCB_byniche_2010, PCB_byniche_2015, PCB_byniche_2020)

```
```{r}

# Sum up fatty acids by fish

# 2010
totalfattyacids_byfish_2010 <- fattyacids_2010_processed %>%
  group_by(Species.ID) %>%
  summarise(Total_Omega3s = sum(Amount),
            Scientific_Name = first(Scientific.Name),
            Common_Name = first(Common.Name),
            Family_Name = first(Family),
            Date = first(Year))

# 2015
totalfattyacids_byfish_2015 <- fattyacids_2015_processed %>%
  # Filter for Omega-3s
  filter(Analyte == "Linoleic acid" | Analyte == 	"Docosahexaenoic acid" |
           Analyte == "Docosapentaenoic acid" | Analyte == "Eicosadienoic acid"
        |  Analyte ==  "Eicosatrienoic acid") %>%
  group_by(Species.ID) %>%
  # Sum up filtered Omega-3s and divide by 1000 to convert from ug/g to mg/g
  summarise(Total_Omega3s = sum(Amount) / 1000,
            Scientific_Name = first(Scientific.Name),
            Common_Name = first(Common.Name),
            Family_Name = first(Family),
            Date = first(Year))

# 2020
totalfattyacids_byfish_2020 <- fattyacids_2020_processed %>%
  # Filter for Omega-3s
   filter(Analyte == "Linoleic acid" | Analyte == 	"Docosahexaenoic acid" |
           Analyte == "Docosapentaenoic acid" | Analyte == "Eicosadienoic acid"
        |  Analyte ==  "Eicosatrienoic acid") %>%
  group_by(Species.ID) %>%
  # Sum up filtered Omega-3s and divide by 1000 to convert from ug/g to mg/g
  summarise(Total_Omega3s = sum(Amount) / 1000,
            Scientific_Name = first(Scientific.Name),
            Common_Name = first(Common.Name),
            Family_Name = first(Family),
            Date = first(Year))

```



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References - i have to make these APA still but just links for now
[1]https://www.nwf.org/Educational-Resources/Wildlife-Guide/Wild-Places/Great-Lakes
<br>
[2] US EPA, O. (2020, June 11). National Coastal Condition Assessment—Great Lakes Human Health Fish Tissue Studies [Overviews and Factsheets]. https://www.epa.gov/choose-fish-and-shellfish-wisely/national-coastal-
condition-assessment-great-lakes-human-health
<br>
[3] https://www.epa.gov/choose-fish-and-shellfish-wisely/2020-great-lakes-human-health-fish-fillet-tissue-study
[4] https://www.epa.gov/system/files/documents/2024-04/glhhffts-qa-fish-sample-preparation-2020.pdf

